ESRC SDAI Application - Understanding domestic abuse using Big Data

### 2019-07-08 case_for_support.tex
Anna update

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will first use the CSEW to investigate the demographic and socio-economic characteristics of survivors of domestic abuse. Exploring the risk factors in domestic abuse victimisation is key in designing effectively targeted policy measures. We will also use the CSEW to explore the police's response to reported cases of domestic abuse. Increasing victim's trust in the police through improving law enforcement response to reported incidents is key in encouraging victims to report domestic abuse and prevent further harm.

@@

We will also use both datasets to explore the predictors of the decision to report and escape the abuse. We are not aware of previous research exploring this question. Using the CSEW and logistic regression approach, we will investigate how the decision to report (or the decision to seek any form of external help) depends on the individual-level, time-invariant characteristics of the survivor, controlling for the severity and type of the abuse suffered. We are especially interested in exploring how trust in the Criminal Justice System (CJS), social ties (living with family, going out, length of living in the area, member of Neighbourhood Watch), financial independence (own income, employment status, car ownership), number of children, mental health, and alcohol and drug dependency affect the decision to report. Using the WMP data, we will explore time-varying factors affecting the decision to leave the abuser. Due to the high levels of underreporting, it is hard to distinguish between time periods when the number of domestic abuse incidents increase, and when the victims have a higher propensity to finally report the ongoing abuse. Here we are mainly interested in the latter case (and will explore the first case in more detail in the last research project), therefore we will focus on the subset of incidents where the victim for the first time reported a long history of abuse. We are interested in whether these reports are more likely to occur after certain days of the year (e.g., birthday of the victim or perpetrator, Christmas, Halloween, Easter, etc.), potentially reflecting cases where the abuse has suddenly escalated, or became known to outsiders.

@@

We will complement this analysis using data on young offenders from the WMP. We will identify if young offenders' home addresses can be linked with a domestic abuse incident in the recent past, allowing us to explore whether living in a household with a domestic abuse problem predict the type of crime they will engage in (property-related, public order offence, violent), controlling for the socio-economic characteristics of the neighbourhood (using the IMD). We are especially interested if living in an abusive environment is predictive of knife crime-related offences.

@@

\textcolor{red}{\textit{Core outcomes}}
 The first UK-based quantitative exploration of the causal effect of domestic abuse and childhood abuse on socio-economic outcomes.
\FloatBarrier
\subsection*{Environmental factors}

\textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will investigate some of the environmental predictors of domestic abuse. While individual-level characteristics are likely to be the main predictors of domestic abuse victimisation, the environment around the victim can play a key role in either exacerbating or alleviating domestic abuse. Understanding how neighbourhood-level characteristics affect the prevalence of domestic abuse and reporting behaviour is crucial for the most effective distribution of resources to help victims and targeted awareness campaigns. We will also explore how time-varying, exogenous factors related to alcohol consumption and financial stability affect domestic abuse.

@@

\textcolor{red}{\textit{Detailed plan}} Using the CSEW, we will first explore whether neighbourhood characteristics are predictive of the prevalence of domestic abuse and the willingness to report. Previous literature has suggested various pathways, including reduced collective efficacy of neighbourhood, and normalisation and acceptance of violence through social norms, to explain the link between neighbourhood characteristics and domestic abuse, but the question has not been explored in a UK context yet. Fisrt, using a logistic regression, we will explore how the interviewer's perception of the street (signs of rubbish, vandalism, and the general condition of houses, member of Neighbourhood Watch) as well as the respondent's connection and attitude towards the neighbourhood (length of time living in the local area, noisy neighbours, rubbish lying around, teenagers hanging around, vandalism, drunken and anti-social behaviour, drug trafficking, abandoned cars, speeding traffic, police presence in the area, worries about crime levels) predict domestic abuse victimisation, after controlling for the socio-economic characteristics of the respondent. Second, we will explore how deprivation in a neighbourhood (LSOA level, measured by the IMD) affects the prevalence of domestic abuse (estimated from the CSEW), and the willingness to report, using a spatial regression framework. Using this statistical approach will alleviate the problem of spatial dependency in our dataset, and will allow us to derive a precise estimate of the effect. These analyses will extend our understanding of the neighbourhood-level predictors of domestic abuse victimisation, and complement our individual-level approach outlined in the first project to provide a comprehensive exploration of the factors affecting domestic abuse victimisation and willingness to report.

We are also interested in exploring how time-varying external factors, including alcohol consumption, and financial stress (measured by gambling expenditure and benefit dependency) affect the propensity to engage in abusive behaviours. In investigating this question, we will use the transaction dataset described earlier, which allows us to estimate a spending profile for each LSOA over time. This will allow us to identify time-specific changes in alcohol consumption, gambling expenses and benefit receipts in parts of the West Midlands, and investigate their effect on the number of reported domestic abuse cases in the area (from the WMP), using a spatial Poisson regression approach.

@@

To explore the effect of financial shocks on domestic abuse, we will be using the same estimation approach, but we will focusing on gambling and benefit payments. Gambling is increasingly recognised as a serious health concern across the UK, and can have a profoundly adverse impact on family finances. We will use our transaction data and WMP dataset to assess the link between changes in spending on gambling in a given LSOA and the prevalence of domestic abuse in that area. The recent reforms of the UK benefit system, in the form of the introduction of the Universal Credit (UC) had been criticised widely, due to the temporal financial strain it imposes on the least financially resilient people in society. Drawing on previous findings about the link between financial stress and domestic abuse, we will explore what effect the roll-out of the UC across the West Midlands had on the reported number of domestic abuse cases. The fact that the exact date of the roll-out varied across the seven metropolitan boroughs within the county allows for a more precise estimation of the effect of UC on the reported number of domestic abuse incidents in the West Midlands.

\begin{table}[!htbp]
\caption{Environmental factors affecting domestic abuse, analysis plan}
  \begin{threeparttable}[t]
  \centering
       \begin{tabular}{ L{2.8cm}  C{1.8cm}  C{1.8cm}  C{3.2cm}  C{2.8cm}  C{2cm} }
    \toprule
     \textbf{Research question} & \textbf{Dataset}    & \textbf{Unit of analysis} & \textbf{Outcome variables} & \textbf{Explanatory variables} & \textbf{Model} \\
    \midrule
    How do environmental characteristics of the neighbourhood predict domestic abuse? & CSEW & Individual (respondent-level) & domestic abuse victimisation/willingness to report & Perception of neighbourhood, socio-economic characteristics & Logistic regression \\
         \midrule
          How does neighbourhood deprivation predict domestic abuse? & IMD, CSEW & LSOA-level & domestic abuse victimisation/willingness to report (proportion of overall population) & Various measures of deprivation & Spatial Logistic regression \\
         \midrule
    How do temporal changes in alcohol consumption and income affect domestic abuse? & WMP, Credit card data & LSOA, day/week-level & Number of reported domestic abuse cases & Spending on alcohol and gambling, changes in benefit receipts & Poisson/Negative binomial regression with spatial effects \\

>>>>

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will first use the CSEW to investigate the demographic and socio-economic characteristics of survivors of domestic abuse. Exploring the risk factors in domestic abuse victimisation is key in designing effectively targeted policy measures. We will also use the CSEW to explore the police's response to reported cases of domestic abuse. Increasing victim's trust in the Criminal Justice System (CJS) through improving law enforcement response to reported incidents is key in encouraging victims to report domestic abuse and prevent further harm.

@@

We will also use both datasets to explore the predictors of the decision to report and escape the abuse. We are not aware of previous research exploring this question. Using the CSEW and logistic regression approach, we will investigate how the decision to report (or the decision to seek any form of external help) depends on the individual-level, time-invariant characteristics of the survivor, controlling for the severity and type of the abuse suffered. We are especially interested in exploring how trust in the CJS, social ties (living with family, going out, length of living in the area, member of Neighbourhood Watch), financial independence (own income, employment status, car ownership), number of children, mental health, and alcohol and drug dependency affect the decision to report. Using the WMP data, we will explore time-varying factors affecting the decision to leave the abuser. Due to the high levels of underreporting, it is hard to distinguish between time periods when the number of domestic abuse incidents increase, and when the victims have a higher propensity to finally report the ongoing abuse. Here we are mainly interested in the latter case (and will explore the first case in more detail in the last research project), therefore we will focus on the subset of incidents where the victim for the first time reported a long history of abuse. We are interested in whether these reports are more likely to occur after certain days of the year (e.g., birthday of the victim or perpetrator, Christmas, Halloween, Easter, etc.), potentially reflecting cases where the abuse has suddenly escalated, or became known to outsiders.

@@

We will complement this analysis using data on young offenders from the WMP. We will identify if young offenders' home addresses can be linked with a domestic abuse incident in the recent past, allowing us to explore whether living in a household with a domestic abuse problem predict the type of crime they will engage in (property-related, public order offence, violent), controlling for the socio-economic characteristics of the neighbourhood (using the IMD). We are especially interested in exploring whether living in an abusive environment is predictive of knife crime-related offences.

@@

\textcolor{red}{\textit{Core outcomes}}
 The first UK-based quantitative exploration of the causal effect of domestic abuse and childhood abuse on socio-economic outcomes.
\FloatBarrier
\subsection*{4. Environmental factors influencing the prevalence of domestic abuse}

\textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will investigate some of the environmental predictors of domestic abuse. While individual-level characteristics are likely to be the most important predictors of domestic abuse victimisation, the environment around the victim can play a key role in either exacerbating, or alleviating domestic abuse. Understanding how neighbourhood-level characteristics affect the prevalence of domestic abuse and the willingness to report is crucial for the most effective distribution of resources to help victims and targeted awareness campaigns. We will also explore how time-varying, exogenous factors related to alcohol consumption and financial stability affect domestic abuse.

@@

\textcolor{red}{\textit{Detailed plan}} Using the CSEW, we will first explore whether neighbourhood characteristics are predictive of the prevalence of domestic abuse and the willingness to report. Previous literature has suggested various pathways, including reduced collective efficacy of neighbourhood, and normalisation and acceptance of violence through social norms, to explain the link between neighbourhood characteristics and domestic abuse, but the question has not been explored in a UK context yet. First, using a logistic regression approach, we will explore how the interviewer's perception of the street (signs of rubbish, vandalism, general condition of houses, Neighbourhood Watch), as well as the respondent's connection and attitude towards the neighbourhood (length of time living in the local area, noisy neighbours, rubbish lying around, teenagers hanging around, vandalism, drunken and anti-social behaviour, drug trafficking, abandoned cars, speeding traffic, police presence in the area, worries about crime levels) predict domestic abuse victimisation, after controlling for the socio-economic characteristics of the respondent. Second, using a spatial regression framework, we will explore how deprivation in a neighbourhood (LSOA level, measured by the IMD) affects the prevalence of domestic abuse (estimated from the CSEW), and the willingness to report. Using this statistical approach will alleviate the problem of spatial dependency in our dataset, and will allow us to derive a precise estimate of the effect. These analyses will extend our understanding of the neighbourhood-level predictors of domestic abuse victimisation, and complement our individual-level approach outlined in the first project to provide a comprehensive exploration of the factors affecting domestic abuse victimisation and willingness to report.

We are also interested in exploring how time-varying external factors, including alcohol consumption, and financial stress (measured by gambling expenditure and benefit dependency) affect the propensity to engage in abusive behaviours. In investigating this question, we will use the transaction dataset described earlier, which allows us to estimate a spending profile for each LSOA over time. This way we can identify time-specific changes in alcohol consumption, gambling expenses and benefit receipts in parts of the West Midlands, and investigate their effect on the number of reported domestic abuse cases in the area (from the WMP), using a spatial Poisson regression approach.

@@

To explore the effect of financial shocks on domestic abuse, we will be using the same estimation approach, but we will focusing on gambling and benefit payments. Gambling is increasingly recognised as a serious health concern across the UK, and can have a profoundly adverse impact on family finances. We will use our transaction data and WMP dataset to assess the link between changes in spending on gambling in a given LSOA, and the prevalence of domestic abuse in that area. The recent reforms of the UK benefit system, in the form of the introduction of the Universal Credit (UC) had been criticised widely, due to the temporal financial strain it imposes on the least financially resilient people in society. Drawing on previous findings about the link between financial stress and domestic abuse, we will explore what effect the roll-out of the UC across the West Midlands had on the reported number of domestic abuse cases. The fact that the exact date of the roll-out varied across the seven metropolitan boroughs within the county allows for a more precise estimation of the effect of UC on the reported number of domestic abuse incidents in the West Midlands.

\begin{table}[!htbp]
\caption{Environmental factors influencing the prevalence of domestic abuse, analysis plan}
  \begin{threeparttable}[t]
  \centering
       \begin{tabular}{ L{2.8cm}  C{1.8cm}  C{1.8cm}  C{3.2cm}  C{2.8cm}  C{2cm} }
    \toprule
     \textbf{Research question} & \textbf{Dataset}    & \textbf{Unit of analysis} & \textbf{Outcome variables} & \textbf{Explanatory variables} & \textbf{Model} \\
    \midrule
    How do environmental characteristics of the neighbourhood predict domestic abuse? & CSEW & Individual (respondent-level) & domestic abuse victimisation/willingness to report & Perception of neighbourhood, socio-economic characteristics & Logistic regression \\
         \midrule
          How does neighbourhood deprivation predict domestic abuse? & IMD, CSEW & LSOA-level & domestic abuse victimisation/willingness to report (proportion of overall population) & Various measures of deprivation & Spatial Logistic regression \\
         \midrule
    How do temporal changes in alcohol consumption and income affect domestic abuse? & WMP, Credit card data & LSOA, day/week-level & Number of reported domestic abuse cases & Spending on alcohol and gambling, changes in benefit receipts & Poisson/Negative binomial regression with spatial effects \\
